An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-18 thesis/compact-finite-difference.tex
fixing more typos

For a single GPU \ref{fig:compact-single-gpu},
when calculating the right hand sides,
each point in the grid is mapped to a single thread,
and each thread applies the required stencil operation
to compute the right hand side at that point.
We note that threads near the left and right boundaries
apply a different stencil from the interior threads.
The implementation of GPU kernels for
stencil operations such as these is a topic of wide study.
The most important considerations were brought out
in the paper by Micikevicius et al. ~\cite{micikevicius20093d}.
For solving the tridiagonal systems,
each thread block of the GPU is mapped to a
different \emph{grid line}
aligned along the
direction the derivatives are being calculated.
The right hand sides are stored
contiguously along these grid lines
(as in Fig. \ref{fig:block-mapping}),
and the modified cyclic reduction algorithm developed
can be used to solve for the derivatives.

>>>>

For a single GPU (Fig. \ref{fig:compact-single-gpu}),
when calculating the right hand sides,
each point in the grid is mapped to a single thread,
and each thread applies the required stencil operation
to compute the right hand side at that point.
We note that threads near the left and right boundaries
apply a different stencil from the interior threads.
The implementation of GPU kernels for
stencil operations such as these is a topic of wide study.
The most important considerations were brought out
in the paper by Micikevicius et al. \cite{micikevicius20093d}.
For solving the tridiagonal systems,
each thread block of the GPU is mapped to a
different \emph{grid line}
aligned along the
direction the derivatives are being calculated.
The right hand sides are stored
contiguously along these grid lines
(as in Fig. \ref{fig:block-mapping}),
and the modified cyclic reduction algorithm developed
can be used to solve for the derivatives.

### 2015-11-18 thesis/introduction-and-background.tex
fixing more typos

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
i.e., the derivative is expressed \emph{implicitly}.
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    f_i^{\prime} + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{dx} + \
    b\frac{f_{i+2} - f_{i-2}}{dx} + \\
    c\frac{f_{i+3} - f_{i-3}}{dx} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints ~\cite{kennedy1994several,lele1992compact},
and $dx$ is the space between two grid points.
We note that for equations of the form in Eq. (\ref{eqn:general-compact}),
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq. (\ref{eqn:general-compact})],
and $A$ is, in general, a \emph{banded} matrix.

@@

\begin{equation}
    f_1^{\prime} + \alpha^{\prime}f_2^{\prime} = \
        \frac{1}{dx}(a^{\prime}f_1 + b^{\prime}f_2 + c^{\prime}f_3 + d^{\prime}f_4)
\label{eqn:boundary-compact}
\end{equation}

>>>>

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
i.e., the derivative is expressed \emph{implicitly}.
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    f_i^{\prime} + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{h} + \
    b\frac{f_{i+2} - f_{i-2}}{h} + \\
    c\frac{f_{i+3} - f_{i-3}}{h} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints ~\cite{kennedy1994several,lele1992compact},
and $h$ is the space between two grid points.
We note that for equations of the form in Eq. (\ref{eqn:general-compact}),
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq. (\ref{eqn:general-compact})],
and $A$ is, in general, a \emph{banded} matrix.

@@

\begin{equation}
    f_1^{\prime} + \alpha^{\prime}f_2^{\prime} = \
        \frac{1}{h}(a^{\prime}f_1 + b^{\prime}f_2 + c^{\prime}f_3 + d^{\prime}f_4)
\label{eqn:boundary-compact}
\end{equation}

### 2015-11-18 thesis/proposed-tridiagonal-algorithm.tex
fixing more typos

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
~\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
We describe an approach for solving such
near-Toeplitz tridiagonal systems efficiently on GPUs.

@@

The forward reduction phase reduces
a $n-by-n$ system of equations to a
$2-by-2$ system of equations in $log_2(n)-1$ steps,
by applying
Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-4})
to every even-indexed equation at each step.
When solving the tridiagonal systems
with the same coefficient matrix,
but repeatedly for different right hand sides,
we note that the results of
Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-3})
remain unchanged for the different right hand sides.
These results correspond to the
coefficients of the tridiagonal system
produced at each forward reduction step.
Thus, given a tridiagonal system,
we may precompute
the coefficients of all
the reduced systems appearing in the
forward reduction steps,
and reuse them for each right-hand side.
We note that for a general tridiagonal system,
this requires storage for $3.(2N - 2)$ coefficients,
in addition to the $3N$ coefficients
for the original tridiagonal system.

Let us consider the case
when the tridiagonal system is \emph{near-Toeplitz},
i.e., when the coefficient matrix is of the form
$A$ in Eq. \ref{eqn:toeplitz-matrix}.
We examine the effect of the
first forward reduction step by making the following substitutions
in Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-3}):

\begin{align*}
    a_2 &= a_3 = a_4 = \hdots &\equiv a_0 \\
    b_2 &= b_3 = b_4 = \hdots &\equiv b_0 \\
    c_2 &= c_3 = c_4 = \hdots &\equiv c_0
\end{align*}
%
We observe that the resulting coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
correspond to the coefficients of a tridiagonal matrix with
exactly the near-Toeplitz structure of $A$.
This form-preserving property of cyclic reduction
has been reported for block Toeplitz tridiagonal systems
by Bini et al. \cite{bini}.

The fact that the reduced system at each step
is near-Toeplitz can be exploited
to reduce the cost of
precomputing and storing the coefficients.
Each near-Toeplitz matrix is completely defined
by only a handful of coefficients:
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
making its storage extremely compact
compared to the case of general tridiagonal systems.
In addition,
it is advantageous to precompute and store the
auxiliary variables $k_1$ and $k_2$,
which can similarly be stored compactly.
With all the forward reduction coefficient matrices
and auxiliary variables precomputed and stored,
the $m$th forward reduction step for equation $i$
is reduced only to the right hand side update:

@@

In this section, we describe the implementation of a GPU solver
for solving a given near-Toeplitz tridiagonal system
for multiple right hand sides.
Our implementation uses
NVIDIA CUDA platform for programming the GPU,
but is easily translated to OpenCL.
The Python programming language is used
to interface with CUDA, by use of the
PyCUDA \cite{kloeckner_pycuda_2012} library.
We develop two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.
The GPU kernels for both implementations are relatively
straightforward and compact,
spanning no more than 100 lines of code.

@@

Figure \ref{fig:cyclic-reduction-precomputing}
shows the
\emph{maximum} amount of storage required for
storing each of the coefficients
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
or auxiliary variables $k_1$ and $k_2$.
A careful analysis of the
forward reduction and backward substitution equations
reveals that the actual amount of storage is somewhat less.
For instance,
the values $b_n^m$ are unused
in Eqs (\ref{eqn:precomputed-forward-reduction-step})-
(\ref{eqn:precomputed-backward-substitution-step-1}).
The values $a_n^m$ and $b_n^m$ are
required only for solving the $2-by-2$
system at the end of the forward reduction phase,
i.e., for $m=log_2(n)-1$,
and they are not stored for the previous steps.
Similarly, the values $c_1^m$ are equal to $c_0^m$,
and do not require special storage.
The set of precomputed coefficients required to be stored is then
$\{a_0^m, b_0^m, c_0^m, k_1^m, k_2^m, b_1^m, k_{1,1}^m, k_{1,n}^m\}$.
Each of these ``coefficient arrays'' are computed on the CPU
and transferred to the GPU.
Additionally, the two scalars $a_n^{log_2(n)-1}$ and $b_n^{log_2(n)-1}$
are required for the 2-by-2 solve.
%
\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/block-mapping.eps}
\end{center}
\caption{Storing right hand sides and mapping to thread blocks.}
\label{fig:block-mapping}
\end{figure}
%
The right hand sides that the system must be solved for
are stored in a single contiguous array in GPU memory
\ref{fig:block-mapping}.
The precomputed coefficient arrays and the right hand side array
are passed as inputs to the compute kernels that implement
the modified cyclic reduction.
We describe two implementations of the kernels,
the first works entirely with global memory,
and the second leverages the GPU's \emph{shared} memory.

@@

The shared memory implementation suffers from two major issues:
first, the number of active threads is halved at each forward reduction step,
(and subsequently doubled at each backward reduction step).
Synchronization between threads of a block is necessary at each step.
Thus, a significant portion of the kernel execution time is
spent by idle threads waiting for active threads to complete execution.
Secondly, the strided access to the
right-hand side values leads to bank-conflicts.
However, the number of bank conflicts is significantly smaller
than in cyclic reduction implementations for general tridiagonal systems,
due the reduced number of computations performed.

>>>>

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
~\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
We describe an approach for solving such
\emph{near-Toeplitz} tridiagonal systems efficiently on GPUs.

@@

The forward reduction phase reduces
a $n-by-n$ system of equations to a
$2-by-2$ system of equations in $log_2(n)-1$ steps,
by applying
Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-4})
to every even-indexed equation at each step.
When solving the tridiagonal systems
with the same coefficient matrix,
but repeatedly for different right hand sides,
we note that the results of
Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-3})
remain unchanged for the different right hand sides.
These results correspond to the
coefficients of the tridiagonal system
produced at each forward reduction step.
Thus, given a tridiagonal system,
we may precompute
the coefficients of all
the reduced systems appearing in the
forward reduction steps,
and reuse them for each right-hand side.
We note that for a general tridiagonal system,
this requires storage for $3.(2n - 2)$ coefficients,
in addition to the $3n$ coefficients
for the original tridiagonal system.

Let us consider the case
when the tridiagonal system is \emph{near-Toeplitz},
i.e., when the coefficient matrix is of the form
$A$ in Eq. (\ref{eqn:toeplitz-matrix}).
We examine the effect of the
first forward reduction step by making the following substitutions
in Eqs. (\ref{eqn:forward-reduction-1}) -
(\ref{eqn:forward-reduction-3}):

\begin{align*}
    a_2 &= a_3 = a_4 = \hdots &\equiv a_0 \\
    b_2 &= b_3 = b_4 = \hdots &\equiv b_0 \\
    c_2 &= c_3 = c_4 = \hdots &\equiv c_0
\end{align*}
%
We observe that the resulting coefficients
$a_i^\prime$, $b_i^\prime$ and $c_i^\prime$
correspond to the coefficients of a tridiagonal matrix with
exactly the \emph{near-Toeplitz} structure of $A$.
This form-preserving property of cyclic reduction
has been reported for block Toeplitz tridiagonal systems
by Bini et al. \cite{bini}.

The fact that the reduced system at each step
is \emph{near-Toeplitz} can be exploited
to reduce the cost of
precomputing and storing the coefficients.
Each \emph{near-Toeplitz} matrix is completely defined
by only a handful of coefficients:
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
making its storage extremely compact
compared to the case of general tridiagonal systems.
In addition,
it is advantageous to precompute and store the
auxiliary variables $k_1$ and $k_2$,
which can similarly be stored compactly.
With all the forward reduction coefficient matrices
and auxiliary variables precomputed and stored,
the $m$th forward reduction step for equation $i$
is reduced only to the right hand side update:

@@

In this section, we describe the implementation of a GPU solver
for solving a given \emph{near-Toeplitz} tridiagonal system
for multiple right hand sides.
Our implementation uses the
NVIDIA CUDA platform for programming the GPU,
but is easily translated to OpenCL.
The Python programming language is used
to interface with CUDA, by use of the
PyCUDA \cite{kloeckner_pycuda_2012} library.
We develop two approaches based on the common idea
of precomputed forward reduction coefficients---one that leverages
the GPU's \emph{shared memory},
and the other working directly with global memory.
The GPU kernels for both implementations are relatively
straightforward and compact,
spanning no more than 100 lines of code.

@@

Figure \ref{fig:cyclic-reduction-precomputing}
shows the
\emph{maximum} amount of storage required for
storing each of the coefficients
$\{b_1, c_1, a_0, b_0, c_0, a_n, b_n\}$,
or auxiliary variables $k_1$ and $k_2$.
A careful analysis of the
forward reduction and backward substitution equations
reveals that the actual amount of storage is somewhat less.
For instance,
the values $b_n^m$ are unused
in Eqs (\ref{eqn:precomputed-forward-reduction-step})-
(\ref{eqn:precomputed-backward-substitution-step-1}).
The values $a_n^m$ and $b_n^m$ are
required only for solving the $2-by-2$
system at the end of the forward reduction phase,
i.e., for $m=log_2(n)-1$,
and they are not stored for the previous steps.
Similarly, the values $c_1^m$ are equal to $c_0^m$,
and do not require separate storage.
The set of precomputed coefficients required to be stored is then
$\{a_0^m, b_0^m, c_0^m, k_1^m, k_2^m, b_1^m, k_{1,1}^m, k_{1,n}^m\}$.
Each of these ``coefficient arrays'' are computed on the CPU
and transferred to the GPU.
Additionally, the two scalars $a_n^{log_2(n)-1}$ and $b_n^{log_2(n)-1}$
are required for the 2-by-2 solve.
%
\begin{figure}
\begin{center}
\includegraphics[height=200pt]{img/block-mapping.eps}
\end{center}
\caption{Storing right hand sides and mapping to thread blocks.}
\label{fig:block-mapping}
\end{figure}
%
The right hand sides that the system must be solved for
are stored in a single contiguous array in GPU memory
\ref{fig:block-mapping}.
The precomputed coefficient arrays and the right hand side array
are passed as inputs to the compute kernels that implement
the modified cyclic reduction.
We describe two implementations of the kernels,
the first works entirely with global memory,
and the second leverages the GPU's \emph{shared} memory.

@@

The shared memory implementation suffers from two major issues:
first, the number of active threads is halved at each forward reduction step,
(and subsequently doubled at each backward reduction step).
Synchronization between threads of a block is necessary at each step.
Thus, a significant portion of the kernel execution time is
spent by idle threads waiting for active threads to complete execution.
Secondly, the strided access to the
right-hand side values leads to bank-conflicts.
However, the number of bank conflicts is significantly smaller
than in cyclic reduction implementations for general tridiagonal systems,
due to the reduced number of computations performed.
